---
title: "lab_02"
author: "derek willis"
date: "2023-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1** Create a new column that divides the expense amounts into the following categories - \$0-\$500, \$500-\$5000, \$5001-\$50000, \$50001\_\$100000, \$100001-\$500000 - giving each an appropriate text value, and count how many expenses are in each group. How would you describe the distribution of expenses in a sentence? Would you change the categories?

**A1.** Most of these expenses are for smaller amounts of less than $500, and very few are larger than $100k. It would make sense to divide the smallest category into several ones to find more detailed differences.

```{r}
maryland_expenses |>
  mutate(
    amount_group = case_when(
      amount <= 500 ~ '0_500',
      amount > 500 & amount <= 5000 ~ '500_5000',
      amount > 5000 & amount <= 50000 ~ '5001_50k',
      amount > 50000 & amount <= 100000 ~ '50k_100k',
      amount > 100000 & amount <= 500000 ~ '100k-500k'
    )
  ) |>
  group_by(amount_group) |>
  summarise(
    total = n()
  ) |>
  arrange(desc(total))
```

**Q2.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A2.** 26 departments have offered at least one class, and Atmospheric and Oceanic Science has the most, with 68.

```{r}
umd_courses |> 
  filter(str_detect(title, "Climate") & seats > 0) |> 
  group_by(department) |> 
  summarize(classes = n()) |> 
  arrange(desc(classes))
```

**Q3.** Of the departments listed in A2, which one is the most interesting to you, and why? **A3.**

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time. **A4.**

```{r}
maryland_expenses |>
  mutate(expenditure_date = mdy(expenditure_date)) |> 
  mutate(month = floor_date(expenditure_date, "month")) |>
  filter(committee_name == 'Moore  Wes For Maryland') |> 
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```

**Q5.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A5.**

```{r}

```
