---
title: "lab_12"
author: "Mitchell Hang"
date: "2023-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

**A1.**  The top word in the list I created was "failing," occurring 1,571 times and accounting for 4.591% of all unique words appearing in the sanctions since 2011. From looking at the list, I can see that many of the words on the list represent the general holding of bank accounts and the transaction(s) of various funds, while "failing" seems to provide a negative connotation, as it could either suggest a loss in funding (or even some form of embezzlement). However, if I add more words to the "stop words" list such as "failing," "client(s)" and "suspension," we now see that the top word is "trust," occurring 534 times and accounting for 1.92% of all of the unique words in the amended list. This leads me to think that many of these lawyers were breaking the trust of their clients by stealing or overly withholding their funds during a trial.

```{r}
md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv")

unique_words <- md_attorney_sanctions |> select(text) |>
  unnest_tokens(word, text)
view(unique_words)

data("stop_words")

unique_words_top_10 <- unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

unique_words_top_10

stop_words_amended <- stop_words |>
  add_row(word = "failing") |> 
  add_row(word = "client") |> 
  add_row(word = "clients") |> 
  add_row(word = "attorney") |> 
  add_row(word = "conduct") |>
  add_row(word = "disbarred") |>
  add_row(word = "suspension") |> 
  add_row(word = "temporary") |> 
  add_row(word = "consent") |> 
  add_row(word = "reprimanded") |>
  add_row(word = "reprimand") |>
  add_row(word = "commission") |>
  add_row(word = "respondent")

unique_words_top_10_amended <- unique_words |>
  anti_join(stop_words_amended) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

unique_words_top_10_amended
```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**A2.** The top trigram is "attorney trust account" and it appears 343 times, accounting for 5.891% of all that was said in the sanctions since 2011. From checking online, an attorney trust account is a type of bank account where client funds are usually kept until a lawyer is able to withdraw them. 

```{r}
md_attorney_sanctions |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  filter(!word3 %in% stop_words$word) |>
  mutate(trigram = paste(word1, word2, word3, sep=" ")) |>
  group_by(trigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

**A3.** The fiscal year which mentions the term "attorney trust account" the most in its sanctions is 2021, with the term appearing 37 times and accounting for 19.577% of all times the term was said in the sanctions since 2011. This leads me to believe that many people had funds in their attorney trust accounts in 2021 as a consequence of the COVID-19 pandemic, and one of the most common infractions for lawyers to commit is stealing escrow funds from their clients' accounts. So it is possible that many lawyers in practice were stealing funds and exploiting those who were disadvantaged (particularly families and businesses) during the pandemic.

```{r}
md_attorney_sanctions |>
  filter(str_detect(text, "attorney trust account")) |>
  group_by(fiscal_year) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100)
```
