---
title: "pre_lab_05.Rmd"
author: "Derek Willis"
date: "2023-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructor demonstration points to hit

1.  Ensure students install refinr package and load it.
2.  Show students how to download and install Open Refine and confirm install before they leave class.

## Chapter 22

### Task 1: Install refinr package

**Task** Run the following code in the gray-colored codeblock below to install refinr.

```{r}
install.packages('refinr')
```

### Task 2: Load libraries and settings

**Task** Run the following code in the gray-colored codeblock below to load the tidyverse, janitor, and refinr and turn off scientific notation.

```{r}
# turn off sci notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
```

### Task 3: Load data

**Task** Run the following code to load that Maryland state government grants and loan data that we've been working with.

```{r}
# Load data
md_grant_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

# Display it
md_grant_loans

```

### Task 4: Group and count to identify problems in Grantee column

**Task** Run the following code to group and count records in the Grantee column, and filter to examine only recipients that start with W. Name a problem that you see that is preventing proper grouping. **Answer**

```{r}
# Now let's try and group and count the number of expenditures by recipient. To make it a bit more manageable, let's use another string function from `stringr` and filter for recipients that start with the uppercase "W" or lowercase "w" using the function `str_detect()` with a regular expression.  

# The filter function in the codeblock below says: look in the Grantee column, and pluck out any value that starts with (the "^" symbol means "starts with") a lowercase "w" OR (the vertical "|", called a pipe, means OR) an uppercase "W". 

md_grant_loans %>%
  group_by(Grantee) %>%
  summarise(
    count=n()
  ) %>%
  filter(str_detect(Grantee, '^w|^W')) %>%
  arrange(Grantee)
```

### Task 5: Use refinr to clean the grantee_clean field.

**Task** Run the following code to make a copy of the Grantee field called grantee_clean, then use the `key_collision_merge` function from refinr to attempt to standardize similar values. Store it as a new object called cleaned_md_grant_loans Click the black arrow on the table to examine the Grantee and grantee_clean columns. Looking at the results on the first 3 pages only (first 30 rows), name two changes the refinr package made, by comparing Grantee to grantee_clean.
**Answer**

```{r}
cleaned_md_grant_loans <- md_grant_loans %>%
  mutate(grantee_clean=key_collision_merge(Grantee)) %>%
  select(Grantee, grantee_clean, everything())

cleaned_md_grant_loans

```

### Task 6: Filter cleaned_md_grant_loans to examine recipients that start with W

**Task** Run the following code to group by grantee_clean and Grantee, then filter for recipients that start with W. Name three changes the refinr function made in the first 30 results. **Answer**

```{r}
cleaned_md_grant_loans %>%
  group_by(Grantee, grantee_clean) %>%
  summarise(
    count=n()
  ) %>%
  filter(str_detect(Grantee, '^w|^W')) %>%
  arrange(Grantee)
```

### Task 7: Use a different refinr method to clean the city_clean field.

**Task** Run the following code to use the n_gram_merge() function to clean the data, then group by grantee_clean and Grantee to examine recipients that start with W. Look at the differences. Does the changes it made vary substantially from the previous use of key_collision_merge? 
**Answer**

```{r}
cleaned_md_grant_loans <- md_grant_loans %>%
  mutate(grantee_clean=n_gram_merge(Grantee)) %>%
  select(Grantee, grantee_clean, everything())

cleaned_md_grant_loans


cleaned_md_grant_loans %>%
  group_by(Grantee, grantee_clean) %>%
  summarise(
    count=n()
  ) %>%
  filter(str_detect(Grantee, '^w|^W')) %>%
  arrange(Grantee)

```

### Task 8: Follow the directions in the textbook to use Open Refine

**Task** Follow the directions in the textbook to upload the Maryland grants & loans CSV to Open Refine (https://thescoop.org/data_journalism_book/open-refine.html#manually-cleaning-data-with-open-refine). Follow the instructions to clean as many recipients as the Open Refine clustering tool can identify. In class, I will come around to look at the Open Refine project you created to do this. When you are finished, export from OpenRefine the cleaned CSV file and put it in the data folder in this pre_lab.
