---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
library(ggrepel)

```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
pg_calls <- read_csv("data/prince_georges_2022_overdoses.csv")

pg_calls <- pg_calls |> mutate(date = date(datetime), month = month(datetime, label=TRUE), week = week(datetime))

```

## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** December accounts for one of every 10 911 overdose calls in Prince George's County, which is not at all like the pattern from Baltimore County. August is near the bottom in PG, while November is near the top. Very different patterns.

```{r}
# Calculate the percentage
pg_calls_month <- pg_calls |>
  group_by(month) |> 
  summarize(total_calls = n()) |> 
  mutate(pct = total_calls/1397 *100) |> 
  arrange(desc(pct))

# Plot the data.  Use geom_line to draw the line, geom_point to add points, geom_text to add labels.  Use scale_x_date to set breaks at 1 day. 
pg_calls_month |>
  ggplot() + 
  geom_bar(aes(x=reorder(month,pct), weight=pct)) + 
  coord_flip() +
  labs(
    title="One in Ten Prince George's Overdose 911 Calls Made in December",
    y = "percentage of calls",
    x = "month",
    caption = "source: Prince George's County EMS"
  )

```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** There's a huge mid-year spike and then a slope upwards towards the end of the year. The sharp decline at the very end of the year is due to four incidents on Dec. 31, the beginning of a new week and the last day of the year.


```{r}
pg_calls_week <- pg_calls |>
  group_by(week) |> 
  summarize(total_calls = n())

pg_calls_week |> 
  ggplot() + 
  geom_line(aes(x=week, y=total_calls)) + 
  labs(
    title="Prince George's County 911 Overdose Calls Show Mid-Year Spike",
    y = "total calls",
    x = "week",
    caption = "source: Prince George's County EMS"
  )

```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  

```{r}
pg_calls_zip <- pg_calls |> 
  group_by(zipcode) |> 
  summarize(total_calls = n())

md_median_income_zip <- get_acs(geography = 'zcta', variables='B19013_001', state='MD', year=2019)

pg_calls_with_income <- pg_calls_zip |> inner_join(md_median_income_zip, by=c('zipcode'='GEOID'))

# Plot it using geom_point.
pg_calls_with_income |>
  ggplot() +
  geom_point(aes(x=total_calls,y=estimate)) +
  geom_text(aes(x=total_calls,y=estimate, label=zipcode), check_overlap = TRUE) +
  labs(
    title="Wealthier Areas of Prince George's Have Fewer 911 Overdose Calls",
    y = "median income",
    x = "number of 911 overdose calls",
    caption = "source: Prince George's County EMS"
  )
```
