---
title: "lab_03"
author: "Derek Willis"
date: "2023-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
md_payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564) |> clean_names()

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |> clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: There are anywhere from a handful to 87 records, depending on how many variations you were able to find.

```{r}
pg_schools <- md_payments |> 
  filter(
    str_detect(vendor_name, "PRINCE GEORGES PUBLIC") | 
    str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") | 
    str_detect(vendor_name, "PG COUNTY PUBLIC") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "BD OF EDUCATION OF PG CO") |
    str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE")
  )
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: Depending on how you did Q1, 2022 had the largest amount, with \$1,578,857,325, and no years should be missing.

```{r}
totals <- pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. There are multiple duplicates in the results, and removing punctuation would be one way to narrow down the list. Another one would be to use mutate with case_when to standardize the vendor names.

```{r}
md_payments |> 
  filter(agency_name == 'CAPITAL PROJECTS- UNIVERSITY OF MARYLAND') |> 
  group_by(vendor_name) |> 
  summarise(total = sum(amount)) |> 
  arrange(vendor_name)
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Starting from a relatively large amount in 2009, stem cell research grants mostly have fallen off since. In 2020, the amount was \$7.1 million.

```{r}
md_grants_loans |> 
  mutate(description = str_to_upper(description)) |> 
  filter(str_detect(description,"STEM CELL RESEARCH GRANT")) |> 
  group_by(fiscal_year) |> 
  summarise(total = sum(amount))
```

Q5. Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5. There are 18 grantees in the data from the 20742 zip code, but several of them are UMCP and could be standardized into a single record for the university. In addition, one of the zip codes is a nine-digit zip and could be made into a five-digit zip code. OVALTZ INC is an interesting one, because I can't find any evidence of it in search results.

```{r}
md_grants_loans |> 
  filter(str_detect(zip_code, '20742')) |> 
  group_by(grantee) |> 
  summarise(total = sum(amount)) |> 
  arrange(desc(total))
```
