---
title: "lab_06"
author: "derek willis"
date: "2022-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)
# load your libraries

```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some of our 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
# load and combine the call data from those three counties
allegany_911<- read_csv('data/allegany_911.csv')
cecil_911<- read_csv('data/cecil_911.csv')
carroll_911<- read_csv('data/carroll_911.csv') %>%
clean_names()
calls_allegany_cecil_carroll<-bind_rows(list(allegany_911, cecil_911, carroll_911))
View(county_voters_combined)
```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore 911 dataframe in creating your join. 

```{r}
# load and join the Baltimore-related data
baltimore_911<- read_csv('data/baltimore_911.csv')
baltimore_911<- baltimore_911 %>% mutate(zip_code = as.character(zip_code))
maryland_zcta<- read_csv('data/maryland_zcta.csv')%>%
clean_names()
maryland_zcta <- maryland_zcta %>% mutate(zcta5n = as.character(zcta5n))
baltimore_911_with_maryland_zcta <- baltimore_911 %>% left_join(maryland_zcta, by=c("zip_code"="zcta5n"))

```

## Answer questions

Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1.**Carroll has the most calls with 448, followed by Cecil with 438 and then Allegany with 418**

```{r}
 calls_allegany_cecil_carroll%>%
  group_by(county) %>%
  summarise(
    count_calls = n() 
  )%>%
arrange(desc(count_calls))
```

Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results. 

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2.**It is noteworthy here that August is the top month for all counties.May also is a high month among all the counties. December is a lower month for both Allegany and Cecil counties. ** 

```{r}
calls_allegany_cecil_carroll_with_month<- calls_allegany_cecil_carroll%>% mutate(month = floor_date(date, "month"))
 calls_allegany_cecil_carroll_with_month%>%
  group_by(county,month) %>%
  summarise(
    count_calls = n() 
  )%>%
arrange(desc(count_calls))
```

Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes. 

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: https://observablehq.com/@palewire/per-capita-calculator.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods, and what else can you tell me about the population there?

A3. **The first zip code is 21217, which covers the  northwestern outskirts of the city. The second zip code is 21223, which covers directly west of the center city. The third zip code is 21202 which covers the eastern center of the city. The fourth zip code is 21213 which covers the northeastern outskirts of the city. The fifth zip code is 21201 which covers the western center of the city. The sixth zip code is 21216 which covers a further west zone from the city. The seventh zip code is 21205 which covers directly east of the city..... The zip code with the highest per capita rate is zip code 21217 which has some neighborhoods in it including Penn North, Harlem Park, Druid Heights, Bolton Hill, Easterwood, Madison Park, Parkview/Woodbrook and Mondawmin. The population in this zip code is 86% black.**

```{r}
baltimore_911_with_maryland_zcta %>% filter(pct_under_18 >= 75)%>%
group_by(zip_code, population) %>%
  summarise(
    count_calls = n() 
  )%>%
mutate (per_capita = count_calls / population * 1000) %>%
  arrange (desc(per_capita))
```

Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4: **I wanted to see how the factor of owning a home plays a role into the number of 911 calls. I found that in the zip code with the most calls that had less then 10% homeowners the population was 9% asian and in the zip code with the most calls that had more then a third of their population owning homes the population was less then 1% asian (.6%)** 

```{r}
baltimore_911_with_maryland_zcta %>% filter(pct_homeowner >= 33)%>%
group_by(zip_code, pct_asian) %>%
  summarise(
    count_calls = n() 
  )%>%
  arrange (desc(count_calls))
baltimore_911_with_maryland_zcta %>% filter(pct_homeowner <= 10)%>%
  group_by(zip_code,pct_asian) %>%
  summarise(
    count_calls = n() 
  )%>%
  arrange (desc(count_calls))
```
```{r}

```
