---
title: "lab_07"
author: "derek willis"
date: "10/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already

```{r echo=FALSE}
census_api_key("YOUR API KEY", overwrite=TRUE)
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS and save them to a dataframe

```{r}
v21 <- load_variables(2021, "acs5", cache=TRUE)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 
**A1** Baltimore City, with an estimated median age of 35.7 years and a margin of error of 0.1 years.

```{r}
md_median_age <- get_acs(geography='county', variables=c(median_age='B01002_001'), state='MD', year=2021)
arrange(md_median_age, estimate)
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 
**A2** Talbot and Worcester counties

```{r}
md_median_age %>% filter(estimate >= 50)
```

# 911 calls and spanish speaking zip codes

**Q3** We're interested in knowing which Maryland counties have the highest percentage of 5-17 year-olds who speak Spanish at home and *also* speak English fluently. Start with the code you wrote for Q1 and alter it to show the total population of 5-17 year-olds by county. Then create a second dataframe to show the number of 5-17 year-olds by county living in homes where Spanish is spoken but who speak English "very well" according to the ACS. To do this, you will need to find the ACS variables that represent those subjects. HINT: the phrase "very well" shows up in multiple variables - you want the one that doesn't include other factors like citizenship.

For each dataframe, rename the `estimate` column to a meaningful name you've given the ACS variable, remove the `variable` columns in each using select and then join them on GEOID and NAME. Then you can add a column to calculate the percentage of the total 5-17 population that speaks Spanish at home.

Which county has the highest percentage of 5-17 year-olds who speak Spanish at home and also speak English very well?

**A3** Prince George's County, with 19%, or nearly one in five.

```{r}
speak_spanish_5_17 <- get_acs(geography='county', variables=c(speak_spanish_5_17='B16004_005'), state='MD', year=2020) %>%
   rename('speak_spanish_5_17'='estimate') %>%
   select(-variable)
md_population_5_17 <- get_acs(geography='county', variables=c(total_pop_5_17='B16004_002'), state='MD', year =2020) %>%
   rename('total_pop_5_17'='estimate') %>%
   select(-variable)

combined_5_17 <- speak_spanish_5_17 %>%
  inner_join(md_population_5_17, by=c('GEOID', 'NAME')) %>%
  mutate(pct_of_total = (speak_spanish_5_17/total_pop_5_17)*100) %>%
  arrange(desc(pct_of_total))
```

**Q4** Given the counties with the largest percentage of 5-17 year-olds that speak Spanish at home, which county stands out to you in the data as different from the others? How would you describe the top 5 counties? How would you describe Garrett County's numbers?

**A4** Talbot County stands out - it has a tiny population compared to Prince George's and Montgomery but as many as 10 percent of its 5-17 year-olds speak Spanish at home and speak English very well. The top 5 are a combination of the largest jurisdictions (except Baltimore city and County) and some very small ones. We may not be able to accurately describe the percentage for Garrett County, given the margin of error, but it seems very, very small.

**Q5** 

**A5** 

```{r}

```

