---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* the tidyverse, janitor and tigris libraries
* the tidycensus library for Q3

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(janitor)
library(tigris)

```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps! 

## Questions 

**Q1.** Maryland voters made history in November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown _based on the total of their respective contests_. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** Moore did better than Brown almost everywhere except Baltimore City and Prince George's County, and in particular did better in rural, whiter counties on the eastern shore. The most interesting county might be Harford.

```{r}
md_statewide <- read_csv("data/md_statewide_county.csv") |> 
  clean_names() |> 
  mutate(geoid = as.character(geoid)) |> 
  mutate(total_ag = peroutka + brown + ag_writein) |> 
  mutate(total_gov = cox + moore + lashar + wallace + harding + sefcik + gov_writein) |>
  mutate(pct_brown = brown/total_ag * 100, pct_moore = moore/total_gov *100, moore_brown_diff = pct_moore - pct_brown)

counties <- counties()

md_counties <- counties |>
  filter(STATEFP == "24")

md_statewide_geometry <- md_counties |> 
  inner_join(md_statewide, by=c('GEOID'='geoid'))

ggplot() +
  geom_sf(data=md_statewide_geometry, aes(fill=moore_brown_diff)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma")

```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.** Talbot customers did not have a good March, with a much higher rate of outages per 10,000 customers than any other county in Maryland. Frederick and Washington counties saw very little power interruption. Given other data, one path would be to create similar maps for other months to see if Talbot was an outlier in March or if there are consistent problems.

```{r}

md_county_outages <- read_csv("data/county_outages_march2023.csv") |> mutate(rate = outage_customers/total_customers*10000)

md_county_outages_geo <- md_counties |> inner_join(md_county_outages, by=c('NAME'='county'))

ggplot() +
  geom_sf(data=md_county_outages_geo, aes(fill=rate)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log")

```

```{r}
outages <- read_csv("~/Downloads/Power_Outages_-_County(1).csv") |> mutate(dt_stamp = mdy_hms(dt_stamp)) |> mutate(month = floor_date(dt_stamp, "month"))
april_2023 <- outages |> group_by(area) |> summarize(total_customers = sum(outages))
```


**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change.What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.** Caroline County stands out for the growth of EVs & hybrids, although it started from a small basis. In general, the bigger percentage change has come in Maryland's rural counties rather than in places where there already was a certain appeal for these vehicles like Montgomery and Baltimore County. The difference between Caroline's change and Dorchester's, right next door, raises some questions in my mind.

For the second map, it's Montgomery and Howard counties, and then every other county in the state, which makes sense. With the exception of Wicomico County, most of the eastern and western parts of Maryland trail far behind.

```{r}
electric_hybrid <- read_csv("data/ev_2020_2023.csv") |> mutate(pct_change = (march_2023-july_2020)/july_2020 *100)

electric_hybrid_with_geo <- md_counties |> inner_join(electric_hybrid, by=c('NAME'='county'))

ggplot() +
  geom_sf(data=electric_hybrid_with_geo, aes(fill=pct_change)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma", trans='log')

md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD")

electric_hybrid_with_geo_pop <- electric_hybrid_with_geo |> inner_join(md_county_population, by=('GEOID')) |> mutate(rate = march_2023/estimate*10000)

ggplot() +
  geom_sf(data=electric_hybrid_with_geo_pop, aes(fill=rate)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma", trans='log')
```


